Viral and Bacterial Communities of Colorectal Cancer

### 2017-06-18 doc/manuscript.Rmd
Added edits from MD.

---
title: Viral and Bacterial Communities of Colorectal Cancer
author:
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
fontsize: 11pt
mainfont: 'Helvetica'
---

@@

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016 and over 500,000 Americans have died from the disease [@Siegel:2014jo]. An important component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb; @Dejea:2014fz; @Arthur:2012kl]. Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association also has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and for identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer. We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous. <!-- I would change the structure of the first two paragraphs to be more linear, but I understand if you want to put it out there as is for time, and because you are probably getting tired of my comments. On high level it reads as "1. Viruses play a role in cancer and evidence points to a role in colorectal cancer. 2. Colorectal cancer is a big deal.  We know about the relationship with microbes. 3. We should study viruses of colorectal cancer." To be less circular, I'd line it up as "1. Colorectal cancer is a big deal. We know about the relationship with microbes. 2. Viruses play a role in cancer and evidence points to a role in colorectal cancer too. 3. We should study viruses of colorectal cancer." -->

Colorectal cancer progresses in a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e., adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus, work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development. <!--I feel this fits oddly here, one solution could be to put these stages of progression/need for early detection above when you introduce why colorectal cancer is a big deal...and how this argument of early detection motivates microbiome research...and now we have the logic to say that virome research could further improve upon this, as we know nothing of the role of viruses...that is the knowledge gap, then leads in well to this last paragraph.-->

@@

We evaluated whether the phages in the community were primarily lytic (obligately lyse their hosts after replication) or temperate (able to integrate into their host's genome to form a lysogen, and subsequently transition to a lytic mode). We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky]. <!-- which of these features was most responsible for identifying the temperate phages here? Guessing #3, but hoping for #1 or #2--> We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus, the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that have the ability to integrate into host bacterial genomes. These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

@@

Here, we have begun to delineate the role the colonic virome plays <!--I think ok to leave this as present tense, as then it reads more generally...we have begun to delineate something general and bigger than this individual study-->in colorectal cancer **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states. By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host <!--ref for this? I know i've read it someone, if not shown explicitely, then in a review, like https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3760236/ and https://www.ncbi.nlm.nih.gov/pubmed/22748808 and https://www.nature.com/articles/srep14802-->.

In addition to modeling the potential connections between virus communities, bacteria communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur. This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu]. We hypothesize that the process began with broadly infectious phages in the colon lysing and thereby disrupting the existing bacterial communities. <!--do we have the time resolution to suggest that the phages started this? and what could trigger that? What about the alternative hypothesis that the microbes began the shift and their phages responded, or (a more specific instance of your proposed case) that integrated prophges responded to cues from an altered environment, e.g., mediated byv a shift in transcriptional regulation of their hosts, that caused them to switch to lysis and drive a community shift. --> This shift led to novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize. Once the initial influential founder bacteria established themselves in the epithelium, secondary opportunistic bacteria were able to adhere to the founders, colonize, and begin establishing a biofilm. Phages may have played a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj]. The oncogenic bacteria may then have been able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response.  As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundances were most influential. As the bacteria entered their oncogenic synergy with the epithelium, we conjecture that the phages continued mediating biofilm dispersal. This process would thereby support the colonized oncogenic bacteria by lysing competing cells and releasing nutrients to other bacteria in the form of cellular lysates. In addition to highlighting the likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future research investigations of the role the virome plays colorectal cancer.

>>>>

---
title: Viral and Bacterial Communities of Colorectal Cancer
author:
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
fontsize: 12pt
mainfont: 'Helvetica'
---

@@

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016 and over 500,000 Americans have died from the disease [@Siegel:2014jo]. An important component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb; @Dejea:2014fz; @Arthur:2012kl]. Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association also has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and for identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer. We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.

Colorectal cancer progresses in a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e., adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus, work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.

@@

We evaluated whether the phages in the community were primarily lytic (obligately lyse their hosts after replication) or temperate (able to integrate into their host's genome to form a lysogen, and subsequently transition to a lytic mode). We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky]. We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus, the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that have the ability to integrate into host bacterial genomes. These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

@@

Here, we have begun to delineate the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states. By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

In addition to modeling the potential connections between virus communities, bacteria communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur. This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu]. We hypothesize that the process began with broadly infectious phages in the colon lysing and thereby disrupting the existing bacterial communities. This shift led to novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize. Once the initial influential founder bacteria established themselves in the epithelium, secondary opportunistic bacteria were able to adhere to the founders, colonize, and begin establishing a biofilm. Phages may have played a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj]. The oncogenic bacteria may then have been able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response.  As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundances were most influential. As the bacteria entered their oncogenic synergy with the epithelium, we conjecture that the phages continued mediating biofilm dispersal. This process would thereby support the colonized oncogenic bacteria by lysing competing cells and releasing nutrients to other bacteria in the form of cellular lysates. In addition to highlighting the likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future research investigations of the role the virome plays colorectal cancer.
